The ChEMBL database as Linked Open Data

### 2012-07-19 article.tex
Reordered the applications, and completed a full draft for the discussion

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable, using extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes the continued project to convert data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies.
\paragraph*{Conclusions:}
The use cases show that by using open standards and ontologies it is possible to link the ChEMBL database
to other databases, such that the integration can understood by software. 
\end{abstract}

@@

The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches~\cite{Ansell2011}. 
A Java Web Archive (WAR) file along with the configuration files and build scripts for the ChEMBL-RDF Linked Data server are available on GitHub~\cite{WebAppGitHub}. 
It proxies the standard ChEMBL-RDF URIs by translating URLs between those requested by users and the URIs that are available
in SPARQL endpoints. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://linkedchemistry.info/chembl/activity/a31863}. If the
user requested an RDF document using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

@@

@@

% Egon,Antony/Valery,Ola: develop and write up the Bioclipse Decision Support use case

@@

\begin{figure*}[!ht]
		\begin{center}
		\includegraphics[width=15cm]{bioclipse-ds.png}
		\newline
		\caption[wee]{Screenshot from Bioclipse Decision Support with results from a ChemSpider + ChEMBL-RDF search. The top left canvas contains the query structure, in this case the drug carbamezapine, the top right canvas shows the near neighbors in ChemSpider (via a similarStructure search) that contain ChEMBL-RDF data, the lower right shows the chemical structure for the selected compound in the top right canvas, and the lower left canvas shows the found interactions for this compound using ChEMBL-RDF.}
	\label{fig:bioclipse-ds}
		\end{center}
\end{figure*}

Chem2Bio2RDF created a single RDF repository covering over 20 public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative search~\cite{Chen2012a} and
mining~\cite{Chen2012b}.  ChEMBL-RDF set (version 13) has been uploaded into Chem2Bio2RDF triple store, enabling to perform
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link each other. For instance,
to investigate the relations between drug side effects and their targets (usually off-targets), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart diseases were linked to 87 unique protein targets (IC$_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein protein interactions by searching across ChEMBL, HPRD and SIDER in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

@@

While we show here that the RDF version of the ChEMBL data is very useful, the current RDF triples are by no means static: ChEMBL-RDF
will keep evolve, following new developments in the Linked Open Data world. For example, it is likely that over time we will link out
to more Linked Data resources. More importantly, we wish to adopt more common ontologies, of which the BioAssay Ontology is planned
to be the next~\cite{Visser2011}. These steps will make the ChEMBL-RDF triples even more interoperable. Part of this future work
is already being performed within the context of the Open PHACTS project~\cite{Williams2012}.

>>>>

\begin{abstract}
\paragraph*{Background:}
Making data available as Linked Data using Resource Description Framework (RDF) promotes integration with other web resources.
RDF documents can natively link to related data, and others can link back using Universal Resource Identifiers (URIs).
RDF makes the data machine-readable, using extensible vocabularies for additional information, making it easier
to scale up inference and data analysis.
\paragraph*{Results:}
This paper describes the continued project to convert data from the ChEMBL database into RDF triples.
This updated version of ChEMBL-RDF now uses recently introduced ontologies, including CHEMINF and CiTO,
exposes more information from the database, and is now available as dereferencable, linked data.
To demonstrate these new features, we present novel use cases showing further integration with
other web resources, including Bio2RDF, Chem2Bio2RDF, and ChemSpider, and showing the use of standard
ontologies.
\paragraph*{Conclusions:}
The use cases show that by using open standards and ontologies it is possible to link the ChEMBL database
to other databases. Using those links data integration can understood by software and create a foundation
for integrative cheminformatics applications, such as the presented decision support.
\end{abstract}

@@

This first applications shows how ChEMBL-RDF can be integrated into other linked data services.
The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches~\cite{Ansell2011}. 
A Java Web Archive (WAR) file along with the configuration files and build scripts for the ChEMBL-RDF Linked Data server are available on GitHub~\cite{WebAppGitHub}. 
It proxies the standard ChEMBL-RDF URIs by translating URLs between those requested by users and the URIs that are available
in SPARQL endpoints. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://linkedchemistry.info/chembl/activity/a31863}. If the
user requested an RDF document using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

@@

A second resource in the linked open data network in which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF created a single RDF repository covering over 20 public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative search~\cite{Chen2012a} and
mining~\cite{Chen2012b}.  ChEMBL-RDF set (version 13) has been uploaded into Chem2Bio2RDF triple store, enabling to perform
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link each other. For instance,
to investigate the relations between drug side effects and their targets (usually off-targets), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart diseases were linked to 87 unique protein targets (IC$_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

@@

% Egon,Antony/Valery,Ola: develop and write up the Bioclipse Decision Support use case
The use of the RDF open standard also simplifies the integration of data resources with 
scientific software, as we have shown previously~\cite{Willighagen2011,Willighagen:2011kx}.
We here demonstrate a new application following this idea.

@@

\begin{figure*}[!ht]
		\begin{center}
		\includegraphics[width=15cm]{bioclipse-ds.png}
		\newline
		\caption[wee]{Screenshot from Bioclipse Decision Support with results from a ChemSpider + ChEMBL-RDF search. The top left canvas contains the query structure, in this case the drug Carbamezapine, the top right canvas shows the near neighbors in ChemSpider (via a similarStructure search) that contain ChEMBL-RDF data, the lower right shows the chemical structure for the selected compound in the top right canvas, and the lower left canvas shows the found interactions for this compound using ChEMBL-RDF.}
	\label{fig:bioclipse-ds}
		\end{center}
\end{figure*}

@@

This papers shows the evolution of the ChEMBL-RDF resource in standardizing access to the
ChEMBL data in a machine-readable manner, by using semantic web technologies. This standardization
applies to several aspects of data exchange, including the data serialization format and
access options, the data structure, and the ontologies used to explain what the data is about.
We touched upon choices made to implement these approaches, such as a selection of
RDF technologies, and chosen ontologies.

For example, for linking to other databases we selected a few predicates to link to
other databases: rdfs:seeAlso, skos:exactMatch, owl:equivalentClass, owl:sameAs. Each
has an increasing level of implications, ranging from "you may also be interested in" to
"both objects are identical and all properties apply to both objects". Additionally,
the CiTO ontology has been used to link data to the literature it was extracted from,
which provides link that lies in between rdfs:seeAlso and skos:exactMatch.

Another ontology that we adopted in this ChEMBL-RDF 13 is the cheminformatics ontology
CHEMINF, allowing us to encode numeric and non-numeric descriptors of chemical entities
in the same standardized way. Using this approach we expose identity descriptors such
as the SMILES string, InChI, and InChIKey, and molecular properties including the
logP, rotatable bond count, and hydrogen bond donor and acceptor counts.

However, not all information is yet using standardized ontologies. For example, the
activity types (IC50, EC50, Potency) are not exposed using an ontology, making it
still impossible to reason or just group over similar activity types. Likewise,
measurement units are provided as literals, making it needlessly hard to compare
concentrations in nM and mM. For this kind of data, two new ontologies were
recently developed. For example, the BioAssay Ontology addresses many of our
remaining concerns around making the experimental data machine readable~\cite{Visser2011},
while the Quantity, Unit, Dimension and Type (QUDT) ontology makes it possible
to express measurements in a machine readable way~\cite{QUDT}. Both approaches are
currently explored to be applied to ChEMBL-RDF as part of work in the
Open PHACTS project~\cite{Williams2012}.

Finally, we stress that nothing in this paper cannot be done with other approaches, such as
more common relational databases. In fact, that is exactly what the ChEMBL website does.
Similarly, there is nothing principle component analysis does that you cannot do with drawings
on paper too; it is just silly. We have here presented a linked data approach to the ChEMBL data
showing standardized and simplified integration to other resources, and that is what
cheminformatics is about.
